Confirmation of Wide-Field Signatures in Redshifted 21~cm Power Spectra using Murchison Widefield Array Observations

### 2015-04-11 ms.tex
Use Paper I instead of Thyagarajan et al.(2015)
From 5b99dfd:
============

In ms.tex,

(*) Change repeated references to Thyagarajan et al. (2015) to Paper I

Our recent study \citep{thy15} used all-sky foreground and instrument models for the first time in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of the cosmological H~{\sc i} power spectrum Fourier modes that are considered most sensitive for detecting the signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

In \citet{thy15}, we verified the general features of our simulations against MWA observations, but were unable to confirm the actual {\it pitchfork} prediction due to insufficient sensitivity in the small amount of data analyzed. Here, we use longer MWA integrations to confirm with high significance the presence of key {\it pitchfork} characteristics of wide-field measurements predicted in the preceding study.

@@

\citet{thy15} have described in detail the effects of wide-field measurements as seen in the delay spectra of interferometer {\it visibilities}. Here, we list the relevant equations and give a brief overview of the wide-field signatures predicted therein.

The delay spectrum for a baseline vector, $\boldsymbol{b}$, is given by \citep{par12a,par12b,thy13,thy15}: 
\begin{align}
  \tilde{V}_b(\tau) &\equiv \int V_b(f)\,W(f)\,e^{i2\pi f\tau}\,\dif f,
\end{align}
with interferometer visibilities, $V_b(f)$, given by \citep{van34,zer38,tho01}:
\begin{align} 
  V_b(f) &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega \\
         &= \iint\limits_\textrm{sky} \frac{A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)}{\sqrt{1-l^2-m^2}}\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif l\,\dif m, 
\end{align}
where, $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, as a function of frequency ($f$) and direction on the sky denoted by the unit vector $\hat{\boldsymbol{s}}\equiv (l,m,n)$, $W_\textrm{i}(f)$ denotes instrumental bandpass weights, $W(f)$ is a spectral weighting function that controls the transfer function in the delay transform, $\dif\Omega=(1-l^2-m^2)^{-1/2}\,\dif l\,\dif m$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector, and $c$ is the speed of light. $\tau=\boldsymbol{b}\cdot\hat{\boldsymbol{s}}/c$ is the geometric delay between antenna pairs measured relative to the zenith and provides a mapping to position on the sky.

The delay power spectrum is defined as \citep{par12a,thy15}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\tilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

@@

Detailed modeling presented in our original study \citep{thy15} has important implications for future instrument design and analysis algorithms:
\begin{enumerate}
\item positioned adjacent to the Fourier modes most sensitive to the cosmological signal, the {\it pitchfork} feature is a leading source of contamination in these modes, and 
\item of the cases considered, a dish-shaped antenna was found to yield the most desirable response from a foreground contamination viewpoint due to a relatively efficient suppression of sensitivity away from the primary field of view. 
\end{enumerate}

@@

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in \citet{thy15} and references therein. In order to reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical beamformer settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots were found to contain amplitude and phase artifacts for a significant duration across different baselines. Hence, they have been excluded from our analysis.

@@

Figure~\ref{fig:delay-spectra} shows the delay spectra obtained from a single snapshot of MWA data (top), averaging LST aligned delay spectra from 12 individual snapshots from MWA observations on different nights (middle), and from modeling (bottom) with no thermal noise fluctuations \citep{thy15} shown for reference. In all panels, the {\it foreground wedge} bounded by horizon limits (white dotted lines) is prominent. The bright branch of power at $\tau=0$ corresponds to foreground emission from the main lobe of the power pattern pointed at the zenith.

@@

The delay power spectra morphologies from data and modeling are remarkably similar in broad shape although they exhibit some differences in the amplitude scales. We attribute these differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in \citet{thy15}.

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. % The models of \citet{thy15} have noted that the foreground power near the horizon limits is due to the nature of wide-field measurements, and is predominantly composed of diffuse emission especially on baseline lengths $\lesssim 100$~m. 
Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in our earlier study.

>>>>

Our recent study \citep[][hereafter referred to as Paper~I]{thy15} used all-sky foreground and instrument models for the first time in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of the cosmological H~{\sc i} power spectrum Fourier modes that are considered most sensitive for detecting the signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

In Paper~I, we verified the general features of our simulations against MWA observations, but were unable to confirm the actual {\it pitchfork} prediction due to insufficient sensitivity in the small amount of data analyzed. Here, we use longer MWA integrations to confirm with high significance the presence of key {\it pitchfork} characteristics of wide-field measurements predicted in the preceding study.

@@

Paper~I describes in detail the effects of wide-field measurements as seen in the delay spectra of interferometer {\it visibilities}. Here, we list the relevant equations and give a brief overview of the wide-field signatures predicted therein.

The delay spectrum for a baseline vector, $\boldsymbol{b}$, is given by \citep[][Paper~I]{par12a,par12b,thy13}: 
\begin{align}
  \tilde{V}_b(\tau) &\equiv \int V_b(f)\,W(f)\,e^{i2\pi f\tau}\,\dif f,
\end{align}
with interferometer visibilities, $V_b(f)$, given by \citep{van34,zer38,tho01}:
\begin{align} 
  V_b(f) &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega \\
         &= \iint\limits_\textrm{sky} \frac{A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)}{\sqrt{1-l^2-m^2}}\,W_\textrm{i}(f)\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif l\,\dif m, 
\end{align}
where, $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, as a function of frequency ($f$) and direction on the sky denoted by the unit vector $\hat{\boldsymbol{s}}\equiv (l,m,n)$, $W_\textrm{i}(f)$ denotes instrumental bandpass weights, $W(f)$ is a spectral weighting function that controls the transfer function in the delay transform, $\dif\Omega=(1-l^2-m^2)^{-1/2}\,\dif l\,\dif m$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector, and $c$ is the speed of light. $\tau=\boldsymbol{b}\cdot\hat{\boldsymbol{s}}/c$ is the geometric delay between antenna pairs measured relative to the zenith and provides a mapping to position on the sky.

The delay power spectrum is defined as \citep[][Paper~I]{par12a}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\tilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

@@

Detailed modeling presented in Paper~I has important implications for future instrument design and analysis algorithms:
\begin{enumerate}
\item positioned adjacent to the Fourier modes most sensitive to the cosmological signal, the {\it pitchfork} feature is a leading source of contamination in these modes, and 
\item of the cases considered, a dish-shaped antenna was found to yield the most desirable response from a foreground contamination viewpoint due to a relatively efficient suppression of sensitivity away from the primary field of view. 
\end{enumerate}

@@

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in Paper~I and references therein. In order to reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical beamformer settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots were found to contain amplitude and phase artifacts for a significant duration across different baselines. Hence, they have been excluded from our analysis.

@@

Figure~\ref{fig:delay-spectra} shows the delay spectra obtained from a single snapshot of MWA data (top), averaging LST aligned delay spectra from 12 individual snapshots from MWA observations on different nights (middle), and from modeling (bottom) with no thermal noise fluctuations (Paper~I) shown for reference. In all panels, the {\it foreground wedge} bounded by horizon limits (white dotted lines) is prominent. The bright branch of power at $\tau=0$ corresponds to foreground emission from the main lobe of the power pattern pointed at the zenith.

@@

The delay power spectra morphologies from data and modeling are remarkably similar in broad shape although they exhibit some differences in the amplitude scales. We attribute these differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I.

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. % The models of Paper~I have noted that the foreground power near the horizon limits is due to the nature of wide-field measurements, and is predominantly composed of diffuse emission especially on baseline lengths $\lesssim 100$~m. 
Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in our earlier study.
